### 2018-02-21 thesis/contents/discussions.tex
Updates from ShareLaTeX

The findings in Study I (Chapter \ref{section:study1}), which aimed to identify trigeminal nerve diffusivity signatures in TN and MS-TN, identified for the first time a method that can distinguish between microstructural changes in these conditions, and revealed localized CN V diffusivity differences in MS-TN and idiopathic TN. The possible implication is that disruption anywhere along both the cisternal and pontine segments of the trigeminal nerve may lead to similar TN-like pain syndromes. Central demyelination in the trigeminal root transition zone is associated with vascular compression in TN. Moreover MS is a classic myelin inflammatory disease. Therefore both diseases may share pathophysiology due to disruptions in myelin ultrastructure. The study opened the possibility that multi-tensor tractograpy aided analysis added a measure of objective assessment in examining pain conditions, and may facilitate similar studies that provide insight into the fundamental changes that occur at the level of white matter.

Study IV (Chapter \ref{section:study4}), which applied group tractography to the study of the trigeminal sensory pathway, further suggested that the unaffected CN V diffusivity is also different from that of controls. DeSouza et al. \cite{Desouza2013} also presented similar findings, where the unaffected RD, MD, and AD were found to differ from controls. Desouza performed the study with manual ROI placement at a single measurement position in the REZ, Whereas we performed GP classification along the nerve automatically. The unaffected side was previously presumed to be similar to normal tissue due to the absence of pain. In many studies it has also been used as a basis of comparison. This finding, however, is not surprising. It has been long observed that bilateral neurovascular compression could be observed in some unilateral TN cases. Moreover, in third-degree neurovascular compression, severe vascular hardening and deformation can be observed in the cisternal space. It is reasonable that these observable macroscopic changes can effect CNS tissue changes on the unaffected side. Moreover descending modulation from PAG can become maladaptive or ineffective in chronic pain, and in these cases more centrally involved changes in the trigeminal nucleus and its associated pathways are possible. With this additional insight, it becomes important to not interpret results of affected-to-unaffected comparisons in terms of differences between affected and healthy controls. Additionally, the logical next question is to determine that, if the unaffected nerve differs from controls, at what degree of difference does it result in pain. Investigations into bilateral TN pain, and vascular compression in the absence of pain, may reveal additional insight into the relationship between white matter changes and TN pain in general.

In Gamma-Knife radiosurgery treatment for TN, the target is on the cistern CN V segment, and situates proximally to the CN V REZ. Paradoxically, the GK treatment can be seen as a willful disruption of the CN V myelination. The question is, why would additional GK lesions on the CN V myelin effectively decrease TN pain. Ephaptic transmission of action potential from the fast conducting fibers to the slow conducting pain fibers is thought to explain TN pain. The paroxysmal nature of TN pain, and the unreliable association between compression and symptom, however, suggest additional factors are involved. Consider that blood circulation and motion is constant within the cistern space, the trigeminal-vascular complex is a dynamically undulating structure. The force exerted onto the CN V from its neighbouring vasculature is therefore periodic. It is possible that during the period where the vascular pressure onto the CN V is at peak, and at the same time the action potential conductance is near the locus of compression, that the ephaptic transmission is triggered and amplified, creating a "resonance condition". This would imply that given periodic vascular forces, and a constant stream of innocuous afferents, the chance of resonance condition is small, and can result in no observable pattern or interval, thus explaining the paroxysmal nature of TN pain. Moreover, under bilateral vascular compression, only the side with vascular and CN V physiology that can result in elevated chance of such resonance condition will result in pain. This model also explains that GK intervention disrupts the chance of resonance condition, thereby reducing TN pain. The pain eventually can return however, as the peripheral lesion site heals and the resonance condition returns. MVD, in this case can be explained as the physical disruption of the condition.

@@

The primary motivation for the study of the trigeminal nerve with tractography was the significant limitation of current conventional MR imaging in the visualization of the brainstem trigeminal nerve and the area of the trigeminal nucleus. Prior to the use of tractography, the structure of the trigeminal nerve could only be inferred from direct anatomical studies or postmortem dissection.

Resolving cross fibers of the cerebral white matter was a fundamental advancement in diffusion tractography. The brain white matter was believed to consists of 90\% complex crossing pathways, and the inability for DTI to reliably image these structure posed a fundamental challenge for neuroimaging. 
A widely adopted technique was probabilistic tractography, which used tissue priors to sample a large number of directions at each step stochastically, and combined the final result as a volumetric probability map. This kind of image-based probabilistic tractography was useful for whole-brain white matter studies. However, it was constrained to the resolution of the diffusion acquisition, which was insufficient for localized and small diameter pathways. Moreover, the algorithm depended on pre-generated tissue-priors, which excluded cranial nerves. 
Deterministic algorithms such as diffusion spectral imaging (DSI), and the related multi-shell acquisition strategies attempted to model the fiber orientation distribution by dramatically increasing the number of gradient directions (256--512 directions were required) in the diffusion acquisition. The acquisition time of these sequences was impractical for the clinical environment. Reduction in acquisition time for these algorithms is currently an active area of research \cite{}.
The family of spherical harmonics (SH) based methods, which include Q-ball and constrained spherical deconvolution (CSD), were a significant development. SH-based methods modelled the fiber orientation distribution as a decomposition of SH basis functions. This meant that the methods are more robust to noise, and therefore require less number of gradient acquisitions. CSD proved to be particularly robust to noise, and can be deployed on acquisitions with only 55 directions. Another family of tractography (XST) algorithm constrained its model assumption to two fiber crossings, also proved to provide good results with 50 gradient directions.

@@

The ability to resolve the pontine trigeminal nerve into the trigeminal nucleus was encouraging. It signified that for the first time we could segment the pontine trigeminal anatomy reliably with computer aid. Previously its identification was done by eye, and sometimes with simple heuristics based on distance from surrounding landmarks. Such manual methods were highly subjective and became problematic with patient images, where distortions to the nerve could not be assessed.

The ability to segment the entire trigeminal nerve also opens the avenue to measure its tissue diffusivity from tractography directly. Moreover, specific tissue-metrics such as myelin image, or PET images, can also be measured with confidence in reproducibility. This extends these methods to facilitate disease studies well beyond TN. We were able to divide the trigeminal nerve into 30 discrete cross-sections along its path, and observed as never before, detailed diffusivity changes as the nerve transits from pontine tissue to the cistern. Determining a common centroid curve from many subjects was computationally non-trivial, especially when the streamline lengths were highly variable, due to the noise in the cistern CSF. Turns out that the merged tractography strategy simplified this task by allowed a single manual curve to be defined manually in the template tractography space. This strategy can similarly be extended for obtaining along-the-track measure for other highly variable streamline bundles.

We applied machine learning to the analysis of TN in Study IV in order to avoid our own confirmation bias. The analysis of the trigeminal nerve in TN versus controls served as validation for the study's method. We had performed traditional multiple comparison statistical analysis, but found the result hard to interpret without inducing biases. The GP classifier was chosen, over more popular SVM classifiers, for the following reasons: 1) the obtained metrics show high location dependency. 2) SVM results are hard to interpret, whereas GP is a bayesian method, and is much more intuitive.

@@

The pain neuromatrix, as studied by various functional and structural connectivity studies, distributes over the full range of neocortical areas. These areas are associated with complex white-matter pathways that extends to all regions of the brain. These include sharp curvatures of the anterior pain pathways of the ACC, the distant projections of the temporal cortices. Some pathways are just beyond the feasibility of current methods, these include the projections of the insula and claustrum, small decending pathways to PAG, and the spinal trigeminal pathways in the brainstem.

We have shown in Study III (\ref{section:study3}), for the first time, that different tractography algorithms are suited to different neuropathways when comparing merged reconstructions in 42 subjects. Maier-Hein et al came to similar conclusions after examining 96 reconstructions from a single set of simulated data \cite{Maier-Hein2017}.
Therefore care must be taken to ensure tractography parameters are targeted to the anatomy of study, as more consistent delineation lead to more robust tissue measurements.

We observed that merged tractography can delineate more complete pathway delineations than from any single subject. This is similar to the observation that averaged MRI anatomical images from multiple subjects can produce high quality templates. Therefore merged tractography may serve to produce high quality tractography templates, that can form the basis for both geometric and measurement comparisons in individuals. We have explored the use of SAGIT in generating the first neonatal porcine tractography template \cite{Zhong2016a}. Such template does not exist for humans to the best of our knowledge, and can be readily created from Human Connectome Project data. In patients where tractography cannot be reliably delineated, for example in the case of severe white matter degeneration in MS or low resolution clinical acquisitions, these template tracts may be used as proxy to enable analysis.

@@

\section{Future development}

The methods developed in this thesis, including SAGIT and merged-tractography GP classifiers,  are not limited to the application of TN. The methods can generalize to any anatomy where consistent tracts can be obtained at a group level. The limitation of its examination power is often in the data, both in data volume, resolution, and quality.

An immediate application is to re-examine TPT tracts, both to establish its diffusivity baseline, and to confirm its asymmetry. This can be immediately deployed on the Human Connectome Project (HCP) dataset. Additional examine in a diseased population requires the availability of clinical MR images that at minimum matches the parameters of HCP.

For pain studies, detailed examination of the pain matrix, such as the medial pain-related white matter projections of the ACC, and posterior insula would be of great interest. Multi-modal and multi-variable study with functional connectivity with this technique would also be possible.

The localization ability of merged-tractography analysis can be readily applied to white-matter lesion studies, and may reveal additional insights in the MS population. Where focal lesion groups can be correlated with disease symptoms such as pain and motor disabilities.

Gaussian Process can be used as a regressor to model tissue parameters along merged white-matter bundles. Any MR scalar image can be embedded into the tractography streamlines, so tissue quantifications such as G-ratio, myelin, proton-density, kurtosis, magnetization transer, PET images can be examined. Since GP is a Bayesian method, the derived GP model can be used to provide predictive confidence on centroid measures from single subjects. When combined with the end-to-end tractography extraction ability of SAGIT, this provides a basis to build a deployable framework for disease inference on new patients.

>>>>

The findings in Study I (Chapter \ref{section:study1}), which aimed to identify trigeminal nerve diffusivity signatures in TN and MS-TN, identified for the first time a method that can distinguish between microstructural changes in these conditions, and revealed localized CN V diffusivity differences in MS-TN and idiopathic TN. The possible implication is that disruption anywhere along both the cisternal and pontine segments of the trigeminal nerve may lead to similar TN-like pain syndromes. Central demyelination in the trigeminal root transition zone is associated with vascular compression in TN. Moreover MS is a classic myelin inflammatory disease. Therefore both diseases may share pathophysiology due to disruptions in myelin ultrastructure. Prior to the study, identification of the CN V nerve inside the brainstem from MR images were highly inconsistent, and the precision of multiple ROI placements could not be maintained. Multi-tensor tractography was instrumental in the feasibility of the method. The study opened the possibility that multi-tensor tractograpy aided analysis added a measure of objective assessment in examining pain conditions, and may facilitate similar studies that provide insight into the fundamental changes that occur at the level of white matter.

Study IV (Chapter \ref{section:study4}), which applied group tractography to the study of the trigeminal sensory pathway, further suggested that the unaffected CN V diffusivity is also different from that of controls. DeSouza et al. \cite{Desouza2013} also presented similar findings, where the unaffected RD, MD, and AD were found to differ from controls. Desouza performed the study with manual ROI placement at a single measurement position in the REZ, whereas we performed GP classification along the nerve automatically. The GP classifier produced results by exhaustively examine moving windows at all position and sizes, and the results are auto-discovered by the algorithm, in contrast to the prior hand-placement of the measurement ROI in previous studies. The consistency in CN V findings in the location and diffusivity across both manual and automatic methods, suggest that the evidences for diffusivity changes in the unaffected CN V are convincing. This finding, however, is not surprising. It has been long observed that bilateral neurovascular compression could be observed in some unilateral TN cases. Moreover, in third-degree neurovascular compression, severe vascular hardening and deformation can be observed in the cisternal space. It is reasonable that these observable macroscopic disturbances can effect CNS tissue changes on the unaffected side. Moreover descending modulation from PAG can become maladaptive or ineffective in chronic pain, and in these cases more centrally involved changes in the trigeminal nucleus and its associated pathways are possible.
The unaffected side was previously presumed to be similar to normal tissue due to the absence of pain. In many studies it has also been used as a basis of comparison. With this additional insight, it becomes important to not interpret results of affected-to-unaffected comparisons in terms of differences between pathological and healthy tissue. Additionally, the logical next question is to determine that, if the unaffected nerve differs from controls, at what degree of difference does it result in pain. Investigations into bilateral TN pain, and vascular compression in the absence of pain, may reveal additional insight into the relationship between white matter changes and TN pain in general.

@@

The primary motivation for the study of the trigeminal nerve with tractography was to overcome the significant limitation of current conventional MR imaging in the visualization of the brainstem trigeminal nerve and the area of the trigeminal nucleus. Prior to the use of tractography, the structure of the trigeminal nerve could only be inferred from direct anatomical studies or postmortem dissections.

The ability to resolve the pontine trigeminal nerve into the trigeminal nucleus was encouraging. It signified that for the first time we could segment the pontine trigeminal anatomy reliably with computer aid. Previously its identification was done by eye, and sometimes with simple heuristics based on distance from surrounding landmarks. Such manual methods were highly subjective and became problematic with the presence of pathology, where distortions to the nerve could not be assessed.

The brainstem CN V nerve in some subjects could be seen to bend caudally into the medulla. Althought not reported in Study IV, sometime we could delineate what appeared to be the spinal trigeminal decussation at the level of the medulla. The fibers were sparse, and not consistently reproducible across our subjects. Moreover, the mesencephalic trigeminal nucleus could be delineated in some subjects as well. Merge tractography demonstrated that these anatomical features have low reproducibility, and therefore they were not included as part of the study. In theory, with enough resolution, these smaller structures can be imaged. It would also be interesting to attempt to delineate the motor branch of the trigeminal nerve.

Resolving cross-fibers of the cerebral white matter was a fundamental advancement in diffusion tractography. The brain white matter was believed to consists of 90\% complex crossing pathways, and the inability for DTI to reliably image these structure posed a fundamental challenge for neuroimaging. 
A widely adopted technique was probabilistic tractography, which used tissue priors to sample a large number of directions at each step stochastically, and combined the final result as a volumetric probability map. This kind of image-based probabilistic tractography was useful for whole-brain white matter studies. However, it was constrained to the resolution of the diffusion acquisition, which was insufficient for localized and small diameter pathways. Moreover, the algorithm depended on pre-generated tissue-priors, which excluded cranial nerves. 
Deterministic algorithms such as diffusion spectral imaging (DSI), and the related multi-shell acquisition strategies attempted to model the fiber orientation distribution by dramatically increasing the number of gradient directions (256--512 directions were required) in the diffusion acquisition. The acquisition time of these sequences was impractical for the clinical environment. Reduction in acquisition time for these algorithms is currently an active area of research \cite{}.
The family of spherical harmonics (SH) based methods, which include Q-ball and constrained spherical deconvolution (CSD), were a significant development. SH-based methods modelled the fiber orientation distribution as a decomposition of SH basis functions. This meant that the methods are more robust to noise, and therefore require less number of gradient acquisitions. CSD proved to be particularly robust to noise, and can be deployed on acquisitions with only 55 directions. Another family of tractography (XST) algorithm constrained its model assumption to two fiber crossings, also proved to provide good results with 50 gradient directions.

@@

The ability to segment the entire trigeminal nerve also opened the avenue to measure its tissue diffusivity from tractography directly. Moreover, this method can extend to any scalar images that can be registered to the DWI space. These include specific tissue-metrics such as myelin, G-ratio, and PET images. This extended these methods to facilitate disease studies well beyond TN. We were able to divide the trigeminal nerve into 30 discrete cross-sections along its path, and observed as never before, detailed diffusivity changes as the nerve transited from pontine tissue to the cistern. The number of divisions chosen was arbitrary, with consideration that the number 30 could easily be divided into 3 sub-segments to represent the beginning, middle, and end of an elongated structure. In theory the number of divisions could be infinitely large, given that a centroid curve was identified. In practice the cross-sectional volume decreased with more divisions, and number of points that were sampled by each cross-sectional volume needed to be considered. Determining a common centroid curve from many subjects was computationally non-trivial, especially when the streamline lengths were highly variable due to the noise in the cistern CSF. White matter pathways are highly complex, with possible multi-directional forks, bends, and convergent pathways across a wide range of scales, some examples include the forniceal crura, zona incerta, and optic chaism. A generalized algorithm that can reliably calculate a centroid for these structure is still a topic of research. Turns out that the merged tractography strategy simplified this task by allowed a single curve to be defined manually in the template tractography space. This greatly reduced the amount of labour for researchers, as centroids no longer need to be defined repeatedly for each subject. Manually curve definition by experts can leverage known techniques in computer graphics to produce centroid for very complex structures, thereby extending SAGIT for obtaining along-the-track measure for other highly variable streamline bundles.

\section{Machine Learning}

We applied machine learning to the analysis of TN in Study IV in order to avoid our own confirmation bias. The analysis of the trigeminal nerve in TN versus controls served as validation for the study's method. We had performed traditional multiple comparison statistical analysis, but found the result hard to interpret. This was largely due to difficulties in determining the right p-threshold, where the choice of p-value can result in profoundly different statistics. Eventually it was decided that frequentist statistics was not suitable to our dataset, as we are burdened with too great amount of data, which contradicts frequentist assumptions. The GP classifier was chosen, over more popular SVM classifiers, for the following reasons: 1) the obtained metrics show high location dependency. 2) SVM results are hard to interpret, whereas GP is a bayesian method, and is much more intuitive. More specifically, the streamline measurements resemble time-series data, in that it exhibits positional dependency between measurements. Unlike a time-series however, the streamline measurements represent physical locations along a pathway, therefore they exhibit bi-directional dependency, as suppose to uni-directional in time. Posing the study as a supervised machine learning problem, the feature vector would be a N dimension vector representing N measures at each division, and a classification label. SVM would assume the features to be points in a N-dimensional space, and each measurement would assume independence. This makes the result of the SVM output difficult to interpret. The GP classifier models the feature vector as a latent GP model, before squashing it to produce a logistic output. Therefore the GP model is able to take the positional dependency into account, and the result output is also much more intuitive. Moreover, changing the GP classification model into a regression one is trivial. This importance is illustrated thus: Study IV asked the question of at which measurement positions can we maximally discriminate TN from controls at a group level, but it is also possible to pose a prognosis question: Given N measurements along a patient's CN V, what is the probability that is resembles a pathological population, what about a healthy population? It is possible to answer these questions trivially with the GP model using regression. The full implication is outside of the scope of this thesis, but its implementation is has already been seeded within.

@@

The findings in S1 in Study IV (\ref{section:study4}) showed TN diffusivity differences in regions surrounding the thalamus, as well as in the mid-segments of the S1 pathways. These findings were in agreement with pain literature, where thalamus and S1/S2 activations are often reported as part of the pain neuromatrix. The TPT pathway resulted in differences near the TGN and the thalamus, areas that are expected to be involved in TN pain. It would be of great interest to re-examine the MS-TN population in these pathways, in order to determine any relationship between deeper white-matter MS-lesion, MS-TN, and non-TN pain in the people with MS.

We have shown in Study III (\ref{section:study3}), for the first time, that different tractography algorithms are differentially suited to delineate different neuropathways when comparing merged reconstructions in 42 subjects. Maier-Hein et al came to similar conclusions after examining 96 reconstructions from a single set of simulated data \cite{Maier-Hein2017}.
Therefore care must be taken to ensure tractography parameters are targeted to the anatomy of study, as more consistent delineation lead to more robust tissue measurements.

Quantification of tissue measures from tractography, as demonstrated in Study IV, implies the use of points in tract models as a sampling method for the scalar values in DWI space. We demonstrated in Study III, that different tratography algorithms exhibit different geometric point distributions. Eventhough we average the point-measures during along-the-tract analysis, which may be considered a form of smoothing/down-sampling. It is not clear whether consistency in comparing measures drawn from the same sampling distribution should be considered as a factor. In other words, given two tracts generatef from identical ROI definitions, with similar cross-sectional area, but of different algorithm, are comparisons between them still valid. Intuitively the algorithm choices and parameters should be identical, and that's what we did in Study III. However, there maybe cases where different parameters are better suited for certain pathologies, and this is may be a concern when deploying tractogrpahy analysis towards clinical diagnosis in the future, and therefore merit further investigation.

@@

\section{Future Directions}

The methods developed in this thesis, including SAGIT and merged-tractography GP classifiers,  are not limited to the application of TN. The methods can generalize to any white matter anatomy where consistent tracts can be obtained at a group level. The limitation of its examination power is often in the data, both in data volume, resolution, and quality.

Extensions to the current studies can follow three lines of investigations:

I) Keep pushing the limit of anatomical delineations towards other areas of the trigeminal system. This include the spinal trigeminal pathways, which include the spinal trigeminal ganglion and its sub-gangions, and its medullary decussations. Re-examine TPT tracts, both to establish its diffusivity baseline, and to confirm its asymmetry. Image the medial pain pathways, and descending projections to the PAG. These tasks are currently limited by the resolution of availabile DWI images. Immediate gains may be obtained by using SAGIT to delineate these structure on the Human Connectome Project (HCP) dataset. Additionally, investigation into 7T DWI images will be necessary.

II) Re-examine TN and other pain in the MS population with S1/S2 and TPT pathways. The localization ability of merged-tractography analysis can be readily applied to white-matter lesion studies, and may reveal additional insights in the MS population. Where focal lesion groups can be correlated with disease symptoms such as pain and motor disabilities. Comparitive studies with other trigeminal pain syndromes such as Herpetic neuralgia and trigeminal deafferentation pain would also be a valuable direction.

III) Continue to advance machine learning to enable disease diagnosis and prognosis from white-matter tractography. Investigate how to diagnose and provide treatment targets from MR images with minimal turn around time. This can involve apply GP regression to the prediction of patient disease, and multi-modal learning, possibly with deep neural networks. Gaussian Process can be used as a regressor to model tissue parameters along merged white-matter bundles. Any MR scalar image can be embedded into the tractography streamlines, so tissue quantifications such as G-ratio, myelin, proton-density, kurtosis, magnetization transer, PET images can be examined. Since GP is a Bayesian method, the derived GP model can be used to provide predictive confidence on centroid measures from single subjects. When combined with the end-to-end tractography extraction ability of SAGIT, this provides a basis to build a deployable framework for disease inference on new patients.

We observed that merged tractography can delineate more complete pathway delineations than from any single subject. This is similar to the observation that averaged MRI anatomical images from multiple subjects can produce high quality templates. Therefore merged tractography may serve to produce high quality tractography templates, that can form the basis for both geometric and measurement comparisons in individuals. We have explored the use of SAGIT in generating the first neonatal porcine tractography template \cite{Zhong2016a}. Such template does not exist for humans to the best of our knowledge, and can be readily created from Human Connectome Project data. In patients where tractography cannot be reliably delineated, for example in the case of severe white matter degeneration in MS or low resolution clinical acquisitions, these template tracts may provide a standard population distribution for GP regression, and can used as proxy to enable advanced analysis.

The pain neuromatrix, as studied by various functional and structural connectivity studies, distributes over the full range of neocortical areas. These areas are associated with complex white-matter pathways that extends to all regions of the brain. These include sharp curvatures of the anterior pain pathways of the ACC, the distant projections of the temporal cortices. Some pathways are just beyond the feasibility of current methods, these include the projections of the insula and claustrum, small decending pathways to PAG, and the spinal trigeminal pathways in the brainstem. Detailed examination of the pain matrix, such as the medial pain-related white matter projections of the ACC, and posterior insula would be of great interest. Multi-modal and multi-variable study with functional connectivity with this technique would also be possible.

From my observations across the studies in TN, I postulate a resonance hypothesis of TN. I am aware that there lacks substantial physiological measurements for these claims, nevertheless I feel it is important to state them here as a possible direction of investigation for the future: In Gamma-Knife radiosurgery treatment for TN, the target is on the cistern CN V segment, and situates proximally to the CN V REZ. Paradoxically, the GK treatment can be seen as a willful disruption of the CN V myelination. The question is, why would additional GK lesions on the CN V myelin effectively decrease TN pain. Ephaptic transmission of action potential from the fast conducting fibers to the slow conducting pain fibers is thought to explain TN pain. The paroxysmal nature of TN pain, and the unreliable association between compression and symptom, however, suggest additional factors are involved. Consider that blood circulation and motion is constant within the cistern space, the trigeminal-vascular complex is a dynamically undulating structure. The force exerted onto the CN V from its neighbouring vasculature is therefore periodic. It is possible that during the period where the vascular pressure onto the CN V is at peak, and at the same time the action potential conductance is near the locus of compression, that the ephaptic transmission is triggered and amplified, creating a "resonance condition". This would imply that given periodic vascular forces, and a constant stream of innocuous afferents, the chance of resonance condition is small, and can result in no observable pattern or interval, thus explaining the paroxysmal nature of TN pain. Moreover, under bilateral vascular compression, only the side with vascular and CN V physiology that can result in elevated chance of such resonance condition will result in pain. This model also explains that GK intervention disrupts the chance of resonance condition, thereby reducing TN pain. The pain eventually can return however, as the peripheral lesion site heals and the resonance condition returns. MVD, in this case can be explained as the physical disruption of the condition.
